---
title: "Project"
author: "Gema Vidal_7526"
date: "March 6, 2017"
output: word_document
---

Loading packages
```{r, eval=FALSE}
library("astsa")
library("forecast")
library("nlme")
```


Setting working directory
```{r, eval=FALSE}
# from Gema's PC
setwd("/Users/gvidal/Box Sync/Vet/MPVM & PhD/PhD 2016_2017 Winter Quarter (gvidal@ucdavis.edu)/STA 137/Project")

# from Gema's Mac
setwd("/Users/gemavidal/Box Sync/Vet/MPVM & PhD/PhD 2016_2017 Winter Quarter (gvidal@ucdavis.edu)/STA 137/Project")

# from Tada's PC
# setwd("INSERT YOUR PATH HERE")

df <- read.table("bostonArmedRobberies.txt", header = FALSE)
head(df)
dim(df)
```




1. Use graphical techniques to inspect the data. Describe the behaviour of the data. Mention any features you think may be important for analysis and forecasting.
```{r, eval=FALSE}
plot(df$V2, type = "l", xlab = "Time", ylab = "Number of Armed Robberies", main = "Number of Monthly Armed Robberies in Boston")
```
Variance: unequal variance. During the initial observations, points are closer together compared with how later in time it tends to spread out.
Trend: increasing trend over time since the number of maximum and minimum number of armed robberies increase over time.
Seasonality:



2. In what way(s) is the data not stationary? Use transformations and/or differencing to make the series stationary. Include a time plot of the transformed and/or differenced data. If you chose a transformation, use this transformation for the remainder steps.

Making the variance equal
```{r, eval=FALSE}
robs_sqrt = round(sqrt(df$V2), 2)
robs_log = round(log(df$V2), 2)
robs_third = (df$V2)^(1/3)
robs_subs = (df$V2)^(-1)

df_trans = data.frame(robs_sqrt, robs_log, robs_third, robs_subs)

# use tada's code that includes the titles of the plots.
par(mfrow = c(2,1))
for (d in df_trans) {
  plot(d, type = "l", xlab = "Time")
}
```


Removing trend by differencing
```{r}
ndiffs(robs_log)
# we have to decide how we want to make the data stationary (transformations? differencing?)
# BASED ON MY PLOTS WITHOUT LINES, I WOULD GO FOR LOG TRANSFORMATION OR POWER TO THE 1/3
# BASED ON FUNCTION ndiffs() WE NEED TO DIFFERENCIATE ONCE TO GET STATIONARY DATA

detrend = diff(robs_log, lag = 1, differences = 1)
ndiffs(detrend)

plot(x = 1:117, detrend, type = "l", xlab = "Time", ylab = "Number of Armed Robberies", main = "Detrended Number of Monthly Armed Robberies in Boston")

Acf(robs_log); Pacf(robs_log)
Acf(diff(robs_log)); Pacf(diff(robs_log))
```


Is the data seasonal? Removing seasonality if necessary
```{r}
nsdiffs(detrend)
```
Gives an error because the data is non seasonal. We don't need to remove seasonality.



3. Using the tranformed and/or differenced data from the previous part, obtain the sample ACF and PACF plots, as well as plots of the raw periodogram, and its smoothed version. Comment on the plots. Use the plots to make a preliminary guess for an appropriate ARIMA model. Keep in mind that differencing plays a part in determining whether the model should be ARMA or ARIMA.


SAMPLE ACF AND PACF
```{r, eval=FALSE}
par(mfrow = c(2, 1))
Acf(detrend, lag.max = 30)
Pacf(detrend, lag.max = 30)

# BASED ON BOTH ACF AND PACF, LOOKS LIKE THEY DECAY AT SIMILAR RATE, SO I THINK IS AN ARMA(P,Q) PROCESS BUT SINCE IT HAS BEEN DIFFERENCED ONCE, THEN I THINK IS AN ARIMA(P, D, Q) WHERE D = 1.
```



RAW PERIODOGRAM AND SMOOTHERED PERIODOGRAM
```{r}
# raw
par(mfrow = c(1,1))
spec.pgram(detrend)

# smoothered (see discussion 9)
spans = c(3, 5, 9, 13, 17, 25, 31)


par(mfrow = c(3,3))
spec.pgram(detrend)
for(k in spans){
spec.pgram(detrend, spans=k, main=paste0('Span ', k))
}
```
Make preliminary guess of appropriate ARIMA / ARMA model would be.




4. Fit the model form the previous step. Include the model, and the parameter estimates. Plot the fitted values and the observed values on the same plot.
```{r, eval=FALSE}
# how to know what model to fit? Looking at ACF and PACF? See discussion 7
# DO NOT USE AUTO.ARIMA() HERE SINCE HE ASKS FOR IT LATER, AND SPECIFIES ARGUMENTS

fit_1 = Arima(robs_log, order = c(1, 1, 1))
fit_1
```
Parameter estimates:


PLOT FITTED VALUES AND OBSERVED VALUES ON SAME PLOT.
```{r}
plot(df$V2, type = "l")
points(x = 1:117, fit_1$fitted, col = 'darkblue', type = 'l')
# IS NOT WORKING, I DON'T KNOW WHY, FITTED POINTS ARE NOT BEING PLOTTED
```




5. Examine the residuals. Provide necessary plots and/or hypothesis test results. Do the residuals resemble Gaussian white noise?
```{r, eval=FALSE}
# residuals
res = fit_1$residuals
res
ts.plot(res)

# sample ACF and PACF
par(mfrow = c(2,1))
Acf(res)
Pacf(res)

# Box-Ljung test (H_0: independence of observations in time series)
Box.test(res, lag=10, type="Ljung")

# histogram and normal QQ-plot
par(mfrow = c(2,1))
hist(res)
qqnorm(res)
qqline(res)
```



6. Use AICc to select an ARIMA model for the (possible transformed) data. Keep in mind that differencing should be incorporated into the model. It is fine to use the function auto.arima() here. It is enough to consider p = 0, ..., 8, q = 0, ...,8, and d = 0, 1, 2. Include the chose model, and provide parameter estimates and their standard errors.
```{r, eval=FALSE}
fit_2 = auto.arima(robs_log, max.p = 8, max.q = 8, max.P = 2, max.Q = 2, max.d = 2, max.D = 1)
fit_2
```



7. Inspect the residuals of this model. Provide necessary plots and/or hypothesis test results. Do the residuals resemble Gaussian white noise?
```{r, eval=FALSE}
# residuals
res = fit_2$residuals
res
ts.plot(res)

# sample ACF and PACF
par(mfrow = c(2,1))
Acf(res)
Pacf(res)

# Box-Ljung test (H_0: independence of observations in time series)
Box.test(res, lag=10, type="Ljung")

# histogram and normal QQ-plot
par(mfrow = c(2,1))
hist(res)
qqnorm(res)
qqline(res)
```



8. Plot the (theoretical) spectral density of the final model together with the smoothed periodogram. Comment on the plots. Describe the method you chose for smoothing the periodogram.
```{r, eval=FALSE}

```



9. Now remove the data for 1975 (the last 10 observations). Using only data from 1966 - 1975 (the first 108 observations) fit an ARIMA model using AICc. Again it is fine to use auto.arima(). Then do the following.
```{r, eval=FALSE}
dim(df)

# taking the first 158 observations.
subset_robs = df[1:108, ]
subset_fsc = df[109:nrow(df), ]
```


a. Write down the chosen model. Include parameter estimates
```{r, eval=FALSE}
fit_3 = auto.arima()

```


b. Inspect the residuals of this model. Do they resemble Gaussian white noise?
```{r, eval=FALSE}
# residuals
res = fit_3$residuals
res
ts.plot(res)

# sample ACF and PACF
par(mfrow = c(2,1))
Acf(res)
Pacf(res)

# Box-Ljung test
Box.test(res, lag=10, type="Ljung") # p-value: 0.5725

# histogram and normal QQ-plot
par(mfrow = c(2,1))
hist(res)
qqnorm(res)
qqline(res)
```


c. Compute point forecasts of the values for January through October 1975. If you used a transformation, then be sure to compute forecasts of the original data, not the transmformed data.
```{r, eval=FALSE}
fst = predict(fit_3, n.ahead = 10)
fst$pred
```


d.Make a time plot of the entire data set, the point forecasts, and 95% prediction intervals. Make another plot of just the observed values from 1975 along with the point forecasts and the prediction intervals. Commnent on the forecast performance.
```{r, eval=FALSE}
# CONFIRM THIS CODE
n = 158
h = 5

par(mfrow = c(1,1))
ts.plot(temp[,2])
polygon(x = c(n+(1:h), n+(h:1)), y = c(upper, lower[h:1]), col = 'lightblue', border = NA)
points(x = n+1:h, y = fst$pred, col = 'purple', type = 'b', pch = 19)

# zoom in
ts.plot(subset_fsc[,2], ylim = c(0.30, 0.48))
points(seq(1,5), fst$pred, col = 'purple', type = 'b', pch = 19)

upper = fst$pred + fst$se
lower = fst$pred - fst$se

polygon(x = c(seq(1,5), seq(5,1)), y = c(upper, lower[h:1]), col = 'lightblue', border = NA)
```

